The effects of Water and Sanitation on Tuberculosis Incidence in South East Asian and African countris

### 2025-03-19 products/manuscript/Manuscript.qmd
Data Analysis (Part 3 of the project)

---
title: "The effects of multi environmental exposures on Tuberculosis Incidence in South East Asian and African countris"
author: ""
format:
  docx:
    toc: false
    number-sections: true
    highlight-style: github
bibliography: ../../assets/dataanalysis-references.bib
csl: ../../assets/american-journal-of-epidemiology.csl
editor: 
  markdown: 
    wrap: 72
---

@@

Tuberculosis is old infectious disease which still exists and become
major public health in the world, particularly in developing countries.
Tuberculosis is linked to environmental factors. This study will analyze
the effects of multi environmental factors on tuberculosis incidence in
Asia and Africa regions, where the majority of tuberculosis coming from
these regions. The environmental factors and country economic and human
development index as covariates.

@@

The study will use multi sources of data including World Bank, World
Health Organization (WHO), United Nation for Children Fund (Unicef),
UN-HABITAT, Emissions Database for Global Atmospheric Research (EDGAR),
and Institute for Health Metrics and Evaluation (IHME). Different
dataset contains different variable, however all dataset have same year,
country name and country codes that can be used to combine the dataset.
The data also contain information of countries globally from 2000 to
2020. The variables will include proportion of urban population living
in slum areas, percentage of population using safely managed sanitation
services, percentage of population using at least basic drinking water
service, fuel for cooking, Population-weighted exposure to ambient PM2.5
pollution, total green houses, and will include GDP/ country economic
status, human development index, and Universal Health Coverage as
covariates.

Note: I have downloaded the dataset, but I have not conducted any
cleaning or exploration on the dataset. List of data sources and website
is provided in an excel fine in my data folder. I will include them in
my reference for the next steps.

@@

The main aim of this study is to look at the impacts of environmental
factors on tuberculosis incidence.

RQ: What environmental factors associated with tuberculosis incidence?

H0: there is no relationship between proportion of urban population
living in slum areas and tuberculosis.

H0: thee is no relationship between safely managed sanitation and
tuberculosis incidence

H0: there is no relationship between drinking water and tuberculosis
incidence

H0: there is no relationship between fuel for cooking and tuberculosis
incidence

H0: there is no relationship between PM 2.5 exposure and tuberculosis
incidence

H0: there is no association between total green gas houses and
tuberculosis incidence

Ha: there is a relationship between proportion of urban population
living in slum areas and tuberculosis.

Ha: there is a relationship between safely managed sanitation and
tuberculosis incidence

Ha: there is a relationship between drinking water and tuberculosis
incidence

Ha: there is a relationship between fuel for cooking and tuberculosis
incidence

Ha: there is a relationship between PM 2.5 exposure and tuberculosis
incidence

Ha: there is a association between total green gas houses and
tuberculosis incidence

@@

The datasets were loded one by one, then drop all unused variable before
comined those datasets using reduce() function. Countries, which have at
least one missing years from 2000 to 2020, were dropped from the study.
To make sure that the data is clean and ready to analyse, all missing
values were dropped. This study focuses on South East Asia and Africa
Regions, where the major proportion of tuberculosis cases came.
Therefore, a newsubset for South East Asia and African countries was
created.

After data cleaning, multiple exploratory data analysis was forformed
such us summary and vizualising the data. A bloxplot was created to show
Tuberculosis Incidence in each country during the time frame. Spagheti
plot was created to visualize the trend of tuberculosis incidence and
other variables over the time. Lastly, line charts showing the trend of
all variables in each country.

The exploratory data analysis showed that the data were not normally
distributed and there was no linear relationship between tuberculosis
incidence and other predictors. Therefore, Spearman's correlation matrix
was perfomed to see the correlations. To have better understanding, a
heatmap matrix plot was created to show the correlation.

The main statistical analysis in thie study is longitudinal data
analysis (eithre fixed effect or mixed effect method). To see which one
is better, Hausman Test was performed to determine which model is
better.

@@

## Data aquisition

Most of the data is downloaded from Worlbank Dataset. However, those
data are originally from many different courses/ organizations, such as
World Bank, WHA, UNICEF, UN-HABITAT, EDGAR, and IHME.

@@

Data Import and data cleaning process have been explained in the methode
part. To see the code for data import and data cleaning , please visit
working directory ("code", "processing-code", "processingfile.qdm").
Processed and clean data are saved at a folder ("data",
"processed-data"). There are many .rds files in the folder icnlduing
file for each variables, and file after combined all variables.

@@

@@

NOTE: I still have issue in saving the stiled tables. I am still working
on it. I will show it next step.

```{r}
#| label: tbl-summarytable
#| tbl-cap: "Data summary table."
#| echo: FALSE
resulttable=readRDS("../../results/tables/summarytable.rds")
knitr::kable(resulttable)
```

The line graph ( @fig-1 ) shows the trend of all variables over the
period in each country. It can be seen that the majority of country
showed the decreases trend in tuberculosis incidence over 21 years
period, except Lesotho showing a sharp increase until the middle of the
period before dramatic decrease until the end of the period. All
countries showed the increase trend in percentage of population using
using basic and safe drinking water, basic and safe sanitation, and
clean fuel and technology for cooking, except Zimbabuwe showing the
decrease in percentage of population using basic drinking water and safe
sanitation, and Senegal showing the decrease of population using clean
fuel and technology for cooking.

@@

( @fig-2 ) shows spaghetti plots in each variables. We can see the
trends of each country in every variables. It can be seen that generally
most of the countries show the increase trends in water and sanitation,
clean fuel use, human development index, total greenhouses, but
fluctuated trends in mean of PM2.5 expousures.

@@

#| label: fig-3
#| fig-cap: "Spaghetti plots in each variables"
#| echo: FALSE
knitr::include_graphics(here("code","eda-code","figures","boxplot.png"))

@@

As first step of statistical analysis. Using scatterplot to show the linearity between tuberculosis incidence and predictors. ( @fig-4 ) shows the scatterplots. The scatterplots do not show linear relationship between tuberculosis Incidence and predictors.

@@

```
Before going further, we need to look at correlation between tuberculosis Incidence and all predictors.

Since they show nonlinearity, spearman's correlation performed and hsowing in the following correlation matrix. See ( @fig-5 )

@@

Since there is no linear relationship, nonlinear relationship test was performed.

@@

For the final model, Generalized linear models (GLMs) and Generalized Addictive Models (GAMs). I need to discuss with Dr. Handal to get advice about this.

I actually have performed both GLMs and GAMs, but I still need to learn both models in depth. Specially for GAM, I am very new in the models. Initially I performed GLMs since the Transformed plots show linears line, but aftrer comparing the GLMs and GAMs model, GAMs perfoming better.

Example @tbl-resulttable2 shows a summary of a linear model fit.

```{r}
#| label: tbl-resulttable2
#| tbl-cap: "Linear model fit table."
#| echo: FALSE
resulttable2 = readRDS(here("results","tables","resulttable2.rds"))
knitr::kable(resulttable2)
```

@@

>>>>

---
title: "The effects of Water and Sanitation on Tuberculosis Incidence in South East Asian and African countris"
author: ""
format:
  docx:
    toc: false
    number-sections: true
    highlight-style: github
bibliography: ../../assets/dataanalysis-references.bib
csl: ../../assets/american-journal-of-epidemiology.csl
editor: 
  markdown: 
    wrap: 72
---

@@

Tuberculosis is old infectious disease which still exists and become
major public health in the world, particularly in developing countries.
Tuberculosis is linked to environmental factors. This study will analyze
the effects of multi environmental factors, specially water and
sanitation on tuberculosis incidence in Asia and Africa regions, where
the majority of tuberculosis coming from these regions. The water and
sanitation will be the main exposures and Human Development Index as a
control variable.

@@

The study will use multi sources of data including World Bank, World
Health Organization (WHO), United Nation for Children Fund (Unicef).
Different dataset contains different variable, however all dataset have
same year, country name and country codes that can be used to combine
the dataset. The data also contain information of countries globally
from 2000 to 2020. The variables will include percentage of population
using safely managed sanitation services, percentage of population using
at least basic drinking water service, percentage of population using
well-managed water, percentage of population using basic sanitation,
percentage of population using well-managed sanitation, and human
development index.

@@

The main aim of this study is to look at the effect of Water and
Sanitation on on tuberculosis incidence.

RQ: Does access to water and sanitation effect Tuberculosis incidence in
Asian and African Regions?

Null Hypotheisis

H0: thee is no effect of access to basic sanitation on tuberculosis
incidence

H0: thee is no effect of access to safely managed sanitation and on
tuberculosis incidence

H0: there is no effect of access to access basic drinking water and
tuberculosis incidence

H0: there is no effect of access to safely managed drinking water on
tuberculosis incidence

Alternative Hypothesis Ha: thee is an effect of access to basic
sanitation on tuberculosis incidence

Ha: thee is an effect of access to safely managed sanitation and on
tuberculosis incidence

Ha: there is an effect of access to access basic drinking water and
tuberculosis incidence

Ha: there is an effect of access to safely managed drinking water on
tuberculosis incidence

@@

The datasets were loaded one by one, then drop all unused variable
before combined those datasets using reduce() function. Countries, which
have at least one missing years from 2000 to 2020, were dropped from the
study. To make sure that the data is clean and ready to analyse, all
missing values were dropped. This study focuses on South East Asia and
Africa Regions, where the major proportion of tuberculosis cases came.
Therefore, a newsubset for South East Asia and African countries was
created.

After data cleaning, multiple exploratory data analysis was performed
such us summary and visualizing the data. A bloxplot was created to show
Tuberculosis Incidence in each country during the time frame. Spaghetti
plot was created to visualize the trend of tuberculosis incidence and
other variables over the time. Lastly, line charts showing the trend of
all variables in each country.

The exploratory data analysis showed that the data were not normally
distributed and there was no linear relationship between tuberculosis
incidence and other predictors. Therefore, Spearman's correlation matrix
was performed to see the correlations. To have better understanding, a
heat map matrix plot was created to show the correlation.

The main statistical analysis in the study is longitudinal data analysis
(either fixed effect or mixed effect method). To see which one is
better, Hausman Test was performed to determine which model is better.

@@

## Data acquisition

The data was downloaded from Worlbank Dataset. However, those data are
originally from many different courses/ organizations, such as World
Bank, WHA, UNICEF, UN-HABITAT, EDGAR, and IHME.

@@

Data Import and data cleaning process have been explained in the method
part. To see the code for data import and data cleaning , please visit
working directory ("code", "processing-code", "processingfile.qdm").
Processed and clean data are saved at a folder ("data",
"processed-data"). There are many .rds files in the folder indulging
file for each variables, and file after combined all variables.

@@

The study applied Generalized Additive Mixed Model (GAMM). The model can
capture mixed effect of countries on Tuberculosis Incidence with non
linear relationship. The data analysis started by looking at the
relationship between Tuberculosis Incidence and Predictors. Non linear
relationship was discovered.Next step is to perform non-linear
relationship to looking at form of relationship. Human Development Index
(HDI), Access to basic water and well-managed water show curvature
relationship, while access to basic sanitation and well-managed
sanitation show polynomial. Data transformation was performed based.
After data transformation, we fit model1, which only includes water and
sanitation variables (as main variables in this study), model2 includes
all predictors (HDI, access to water and sanitation), model3 only
includes significant variables in the previous model. Model 1, 2 and 3
use Generalized Linear Mixed Model (GLMM). We also performed VIF test to
look at multicollinearity. Model 4 uses Generalized Additive Mixed
Models. Moreover, BIC and AIC tests were performed to look at the best
model. BIC and AIC tests found that Model 4 (GAMM) performs better with
the smallest BIC and AIC. To be able to interpret GAMM better, we
visualize the result into plots.

@@

NOTE: I still have issue in saving the tables. I am still working on it.
I will show it next step.

```{r}
#| label: tbl-summarytable
#| tbl-cap: "Data summary table."
#| echo: FALSE
resulttable=readRDS (here("code", "eda-code", "tables", "summary_table.rds"))
knitr::kable(resulttable)
```

The line graph ( @fig-1 ) shows the trend of all variables over the
period in each country. It can be seen that the majority of country
showed the decreases trend in tuberculosis incidence over 21 years
period. All countries showed the increase trend in percentage of
population using using basic and safe drinking water, basic and safe
sanitation, except Zimbabuwe showing the decrease in percentage of
population using basic drinking water and safe sanitation.

@@

( @fig-2 ) shows spaghetti plots in each variables. We can see the
trends of each country in every variables. It can be seen that generally
most of the countries show the increase trends in water and sanitation,
and human development index

@@

#| label: fig-3
#| fig-cap: "Boxplot plots for each Country"
#| echo: FALSE
knitr::include_graphics(here("code","eda-code","figures","boxplot.png"))

@@

As first step of statistical analysis. Using scatter plot to show the
linearity between tuberculosis incidence and predictors. ( @fig-4 )
shows the scatter plots. The scatter plots do not show linear
relationship between tuberculosis Incidence and predictors.

@@

Before going further, we need to look at correlation between
tuberculosis Incidence and all predictors.

Since they show non linearity, spearman's correlation performed and
showing in the following correlation matrix. See ( @fig-5 )

@@

Now, we want to assess non linear relationship.

@@

There are four models build; fitting model with only water and
sanitation variables, then fitting models including co-variate (HDI).
The third model is to fit only significant variables. The first three
models use Generalized Linear Mixed Models (GLMM), while the last model
used Generalized Additive Mixed Models (GAMM). Those models were
compared using AIC and BIC criteria to select the best/ the final
models.

```{r}
#| label: tbl-summarytable
#| tbl-cap: "Model Criteria"
#| echo: FALSE
resulttable2=readRDS (here("code", "analysis-code", "tables", "criteria_output.rds"))
knitr::kable(resulttable2)
```

Based on BIC and AIC criteria, the best model fit the data is
Generalized Additive Mixed Models.

\$ \log(\text{TBI}) \sim s(\text{hdi}) + s(\text{pro\_basicwater}) +
s(\text{pro\_safewater}) + s(\text{pro\_basicsanitation}) +
s(\text{country}, \text{bs}="re") \$

```{r}
#| label: tbl-summarytable
#| tbl-cap: "Generalized Additive Mixed Models"
#| echo: FALSE
resulttable3=readRDS (here("code", "analysis-code", "tables", "summary_Gamm.rds"))
knitr::kable(resulttable3)
```

This study employed a Generalized Additive Model (GAM) with a Gaussian
family and identity link function to examine the association between
tuberculosis incidence and key environmental and socioeconomic factors
across multiple countries. The model incorporated smooth functions for
continuous predictors and a random effect for country to account for
within-country variability. The results indicate that HDI and basic
sanitation are significant determinants of tuberculosis incidence, while
access to basic and safe water did not show strong associations.

The smooth function for Human Development Index (HDI) was highly
significant (p \< 2e-16), with an estimated effective degrees of freedom
(EDF) = 4.283, confirming a non-linear relationship between HDI and
tuberculosis incidence. The effect was strongest at lower HDI values,
suggesting that improvements in socioeconomic conditions may have the
greatest impact in reducing tuberculosis burden in low-HDI countries.

Among the water and sanitation variables, basic water access was not
statistically significant (p = 0.73379), indicating no strong evidence
of an association between water coverage and tuberculosis incidence.
Similarly, safe water access was not significant at α = 0.05 (p =
0.05605), suggesting only a weak relationship. However, basic sanitation
access was statistically significant (p = 0.00542, EDF = 3.845),
suggesting that higher access to sanitation is associated with a
reduction in tuberculosis incidence. This highlights the importance of
improving sanitation infrastructure as a potential intervention for
tuberculosis control.

The random effect of country (s(country)) was highly significant (p \<
2e-16, EDF = 16.768), indicating substantial between-country variation
in tuberculosis incidence. This justifies the inclusion of random
effects to account for country-level heterogeneity and underscores the
need for context-specific policy interventions tailored to national
conditions.

The model fit was strong, with an adjusted R² of 0.974, indicating that
97.4% of the variance in tuberculosis incidence was explained by the
model. The scale estimate was 0.014598, suggesting a low residual
variance, which reflects a well-fitting model.

```{r}
#| label: fig-7
#| fig-cap: "Generalized Additive Mixed Models"
#| echo: FALSE
knitr::include_graphics(here("code","analysis-code","figures","gam_diagnostics.jpg"))

@@

These findings underscore the significant role of socioeconomic
conditions and sanitation infrastructure in tuberculosis control. While
access to basic and safe water was not a significant predictor,
improvements in sanitation and overall socioeconomic development (HDI)
were strongly associated with lower tuberculosis incidence. Given the
significant country-level differences, tuberculosis prevention
strategies should be tailored to national contexts, emphasizing
improvements in living standards and sanitation access. These results
provide valuable insights for policymakers aiming to design targeted
public health interventions in high-burden settings.
